---
title: "fortedata: Working with the data"
author: "Jeff Atkins"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fortedata: Working with the data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  require(fortedata),
  require(ggplot2),
<<<<<<< HEAD
  require(viridis),
=======
  require(viridis), 
>>>>>>> 35aec8de476b33311f789ac602db0036d3348810
  require(magrittr),
  require(dplyr)
)

```

## Working with `fortedata` 
In order to help end-users work with `fortedata` we have included some worked examples.


## Example No. 1:  Working with disturbance and treatment assignments


As detailed in the `fd_experimental_design_vignette`, FoRTE's stem-girdling disturbance has four severity levels--0, 45, 65, and 85% targeted leaf area reductions--applied in two ways, a bottom-up and top-down treatment. The plot metadata and assignments are contained in `plot_metadata`.

For data analysis it is often necessary to match disturbance metadata with other FoRTE data. For our example, we are going to look at how hemispherical camera derived leaf area varies from pre- (2018) to post-disturbance (2019), by disturbance severity level and treatment.

First, we want to bring the metadata into the workspace

```{r disturb1}

# bring in metadata via the plot_metadata() function
df <- fortedata::fd_plot_metadata()

# now we converte the tibble to a data frame
df <- data.frame(df)
```

If we look at the structure of the disturbance metadata with `str` we can see that the metadata includes plot information for all plot hierarchies within the FoRTE design, including replicate, plot, subplot, and nested subplot.

```{r disturb-2}
# let's look at the structure of the data
str(df)
```

For our example analysis we want to isolate the `subplot_id` level metadata to match with our camera data.

```{r disturb-3}
# First we want to concatenate our replicate, plot and subplot data to make a subplot_id column 
df$subplot_id <- paste(df$replicate, 0, df$plot, df$subplot, sep = "")
df$subplot_id <- as.factor(df$subplot_id)

# Now that we have our data in the form for this analysis, let's filter our metadata to the subplot level.
df %>%
  dplyr::select(subplot_id, disturbance_severity, treatment) %>%
  dplyr::distinct() %>%
  data.frame() -> dis.meta.data

# this filters the metadata down to the subplot_id level
dis.meta.data <- dis.meta.data[c(1:32), ]

print(dis.meta.data)
```

Now that we have our metadata in the right format, we want to bring in our camera data using the `fd_hemi_camera()` function and then merge with our metadata.

```{r disturb-4}
# First we import the camera data
cam <- fd_hemi_camera()

# Then we merge with the metadata from above
cam <- merge(cam, dis.meta.data)

# For this analysis we want to code both disturbance severity and treatment as factors
cam$disturbance_severity <- as.factor(cam$disturbance_severity)
cam$treatment <- as.factor(cam$treatment)
```

Now we to analyze the data for differences. First, let's plot our data to see what we are dealing with. We want to look differences from pre- to post-disturbance by both disturbance severity and treatment. First, we will make a new column by extracting the year from the `date` column then we will do a facetted density plot to look at distributions. 

```{r disturb-5, fig.align = "center", fig.width = 6, fig.height = 4}
# making year column
cam$year <- format(as.Date(cam$date), "%Y")
# let's look at distribution by year, by disturbance severity and treatment
ggplot(cam, aes(x = lai_cam, fill = disturbance_severity ))+
  geom_density()+
  xlab("Leaf Area Index")+
  ylab("")+
  facet_grid(year ~ treatment)

```

This is a pretty good first assumption, but let's clean it up. FoRTE uses a standardized color scheme that can be accessed via `forte_colors()`. The palette is such that the control is black ("#000000"), 45% is green ("#009E73"), 65% is blue ("#0072B2"), and 85% is orange ("#D55E00"). The specfic hex codes are used as they are eye-friendly color palettes. 


```{r colors, fig.align = "center", fig.width = 6, fig.height = 4}
forte_pal <- forte_colors()

# let's make a function to print the palette
print.palette <- function(x, ...) {
  n <- length(x)
  old <- par(mar = c(0.5, 0.5, 0.5, 0.5))
  on.exit(par(old))

  image(1:n, 1, as.matrix(1:n), col = x,
        ylab = "", xaxt = "n", yaxt = "n", bty = "n")

  rect(0, 0.9, n + 1, 1.1, col = rgb(1, 1, 1, 0.8), border = NA)
  text((n + 1) / 2, 1, labels = attr(x, "name"), cex = 1, family = "serif")
}

print.palette(forte_pal)
```
We can then use the colors from `forte_colors` and create a cleaner plot with a few extra lines of code. 

```{r disturb-6, fig.align = "center", fig.width = 6, fig.height = 4}
# first let's make some new, more informative labels for our facets
facet.labs <- c("B" = "Bottom-Up", "T" = "Top-Down")

# let's look at distribution by year, by disturbance severity and treatment
ggplot(cam, aes(x = lai_cam, fill = disturbance_severity ))+
  geom_density(alpha = 0.4)+
  theme_minimal()+
  scale_fill_manual(values = forte_pal,
                    name = "Disturbance Severity",
                    labels = c("0%", "45%", "65%", "85%"))+
  theme(legend.position = "bottom")+
  xlab("Leaf Area Index")+
  ylab("Density")+
  facet_grid(year ~ treatment, labeller = labeller(treatment = facet.labs))
```
