---
title: "fortedata: Forest Inventory Data"
author: "Jeff Atkins"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fortedata: Forest Inventory Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  require(fortedata),
  require(ggplot2),
  require(viridis)
)
```

## Forest Inventory Data
The `fd_inventory()` script within `fortedata` currently includes one function:

* `fd_inventory()` returns a single dataset of the forest inventory data, including diameter-at-breast height (DBH), latitude, longitude, and biomass for each measured stem, as well as information on vitality and canopy position . There are 3165 observations in the dataset, all measured in 2018.

```{r fd_inventory}
fd_inventory()
```
## Inventory Data
  
  Biomass can be calculated from `fd_inventory()` directly, without assignment, by using the function `calc_biomass()` as such using site/regionally specific allometries (details need to be added.)

```{r biomass, fig.height = 3.5, fig.width = 7.5, fig.align = "center"}
# import the biomass data
df <- fortedata::calc_biomass()

# return density plot of replicate biomass
ggplot2::ggplot(df, ggplot2::aes(x = biomass, fill = as.factor(replicate)))+
  ggplot2::geom_density(alpha = 0.4)+
  ggplot2::theme(legend.position = "none")+
  ggplot2::facet_grid(.~replicate)
```
