---
title: "fortedata: Remote Sensing Data"
author: "Jeff Atkins"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FoRTE Remote Sensing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  require(fortedata)
)
```

## Remote Sensing Data

The `fd_remote_sensing()` script within `fortedata` currently includes the following functions:

* `fd_hemi_camera()` returns a single dataset that includes derived estimates of leaf area index, gap fraction, clumping index, and NDVI (normalized difference vegetation index) from terrestrial, upward-facing hemispherical photos looking into the forest canopy taken 1 meter above-ground (Table S9). The data package includes 1028 observations of each variable from 2018 and 2019 (Fig. 4).

* `fd_canopy_structure()` returns a single dataset that includes estimates of canopy structural traits such as height, area/density, openness, complexity, and arrangement derived from terrestrial lidar and processed using  forestr version 1.0.1 (Atkins et al. 2018) in R Version 3.6.2 (R Core Team, 2020). The package includes 62 observations for each metric (28 canopy structural metrics are included in forestr v1.0.1 that estimate canopy structural traits such as area/density, openness, arrangement, heterogeneity, and layering (Atkins et al. 2018; Fahey et al. 2019)) from 2018 (Table S10). 

* `fd_ceptometer()` returns a single dataset that includes estimates of the fraction of photosynthetically available radiation (faPAR) absorbed by the canopy as well as leaf area index (LAI_cept)--each derived from a handheld ceptometer (LP-80; Decagon Devices) (Table S11). The package includes 32 observations of each variable from 2019 and 16 from 2018 (Fig. 4). 


## Hemispherical Camera Imagery
  
Below-canopy, hemispherical images  were taken using a 24 Megapixel DSLR camera (Regent Instruments; Quebec, QU, Canda) with a 180° hemispherical lens during peak leaf-out (~July). The camera was facing-upwards, into the canopy and was mounted on a self-leveling tripod with the lens at 1 m from the ground. Leaf area index (LAI) was estimated using WinSCANOPY (Regent Instruments).

```{r fd_hemi_camera}
fd_hemi_camera()
```

## Canopy Structural Traits (CSTs)

We derived canopy structural trait metrics (Fahey et al. 2019) from 2-D terrestrial lidar data collected with a user-mounted, portable canopy lidar (PCL) system equipped with an upward facing, Riegl 3100VHS near‐infrared pulsed‐laser operating at 2000 Hz (Riegl LD90 3100 VHS; Riegl USA Inc., Orlando, Florida, USA). The PCL generates continuous LiDAR returns from a “slice” of the canopy as it is walked along a measured transect. For this study, we used 40 m transects at cardinal directions--north-to-south,  east-to-west--through subplot center for a total of 80 m of transect length per subplot. This mirrors the approach of Atkins et al. (2018) and Haridman et al. (2013) and is sufficient to account for spatial variability of forest structure (Hardiman et al. 2019).  A more detailed description of the development and implementation of this terrestrial laser scanning system is available in Parker et al. (2004) and Hardiman et al. ( 2011).  We derived canopy structural metrics using the forestr package (Atkins et al. 2018a, b) in R 3.5 (R Core Team, 2018). forestr produces a suite of canopy structural traits metrics that describe canopy area/density, openness, height, complexity, and arrangement (Atkins et al. (2018a). 

```{r fd_canopy_structure}
fd_canopy_structure()
```

## Light Interception and Availbility

```{r fd_ceptometer}
fd_ceptometer()
```
